Omics-based Single Step Trait Prediction

### 2017-09-12 paper/manuscript.tex
paper, change 'Core' to 'Reduced'

\section{Materials and Methods}
\Genetics2level{Exp1}
To address objectives 2 and 3, a maize inbred line data set, subsequently
referred to as \textit{Exp1}, was used~\cite{Yang2014}.
\textit{Exp1} originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploting identify by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modelling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp2}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 parental Dent and 43 parental Flint lines, that are parents of 685
hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

@@

\Genetics2level{Population structure}
Population structure in \textit{Exp1} was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the R-pacakge LEA
\cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \textit{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

@@

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the $l$-th group of inbred lines 
($\mathbf{W}_{l}$) has dimensions $n_{l} \times p$ where $n_{l}$ pertains to 
the number of genotypes in the $l$-th group and $p$ pertains to the number of 
features.
Note that $l$ is only defined for \textit{Exp2} where a distinction between 
the two heterotic groups is necessary whereas $l$ can be ignored for all 
models pertaining to the inbred lines of the maize diversity panel in 
\textit{Exp1}.
To avoid mentioning special cases, we will use the general notation with 
the index $l$ throughout this section, though.
For the diversity panel maize lines, $n_{l}$ corresponds to the number of 
genotypes whereas, in the case of the hybrid data, $n_{l}$ corresponds to the 
number of parent lines in the $l$-th heterotic group.
In generating kernels for each predictor, all features in $\mathbf{W}_{l}$ were
centered and standardized to unit variance.
Then, each kernel can be defined as

@@

Following the notation in \citeNP{Fernando2014} the phenotypes can be untangled 
like this:

@@

\subsection{Predictive ability and model validation}
For the validation of our predictions, we employed leave-one-out
cross-validation (LOOCV) routines.
In the case of the diversity panel maize inbred lines, LOOCV was performed by 
using a single genotype as a hold-out sample, which will subsequently be 
predicted, while using all other inbred lines for model training.
This process is repeated until all inbred lines have been used once for testing
and $n - 1$ times for model training.

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred.pdf}
  \caption{
  Predictive abilities and SE for tropical/subtropical inbred lines from the 
  maize diversity panel and six agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  The `Core' panel includes a subset of 110 lines covered by genomic and
  transcriptomic information whereas the `Full' panel comprises all 164
  genotypes.
  In the `Full' set of genotypes, transcriptomic records were imputed for 54
  lines.
  }
\label{fig:InbredResults}
\end{figure}

\Genetics2level{Exp2}
The maize hybrid data set contains information on seven agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Core'.
In the `core' set, transcriptomic data were the best single predictor for ADL,
PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC\@.
For the FAT and SUG, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/pred_ability_hybrid.pdf}
  \caption{
    Predictive abilities and average, bootstrapped standard errors for the set
    of maize hybrids and six agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The `Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the `Core' set comprises 685 hybrid progeny
    based on 103 parental inbred lines.
  }
\label{fig:HybridResults}
\end{figure}

>>>>

\section{Materials and Methods}
\Genetics2level{Exp1}
To address objectives 2 and 3, a maize inbred line data set, subsequently
referred to as \textit{Exp1}, was used~\cite{Yang2014}.
\textit{Exp1} originally comprised 513 maize lines representing the global maize
diversity  and was reduced to the set of tropical and subtropical lines
($n = 211$) given that it is the largest of the four pre-classified
subgroups~\cite{Yang2014}.
All inbred lines were evaluated in five Chinese environments ranging from
$18$ to $30^{\circ}$N and from $102$ to $110^{\circ}$E~\cite{Yang2014}.
Best linear unbiased predictors (BLUPs) were calculated for 17 traits of which 
six are presented in this study (Table~\ref{table:TraitKey}).
The selection of these traits was done so that four traits were better 
predicted by transcriptomic data and two further traits were better predicted 
by genomic data.
MaizeSNP50 BeadChip-data were available for the entire set of lines from the
diversity panel.
Additionally, RNA sequencing was performed for 368 lines on young seedlings 15 
days after sowing.
By exploiting identity by descent (IBD) between 49,728 SNPs on the BeadChip 
overlapping with the RNA-seq data, 556,809 high quality SNPs could be inferred 
for these lines~\cite{Fu2013,Li2013}.
For the remaining 145 maize lines without RNA-seq data, high density markers
were inferred via projection of IBD regions onto the BeadChip, using this core 
set of markers~\cite{Yang2014}.
The same SNP quality checks as for the parental inbred lines of the hybrids
were applied to the 211 inbred lines from the tropical/subtropical subset,
yielding 37,760 SNPs.
Gene expression data were normalized using a normal quantile transformation to
satisfy modeling assumptions~\cite{Fu2013}.
A subset of 28,850 annotated genes was available for 149 out of the 211
genotypes and was kept for further analyses~\cite{Li2013}.

\Genetics2level{Exp2}
To address objectives 1 and 2, a maize hybrid data set, subsequently
denoted as \textit{Exp2}, was used.
The material comprised 1,521 hybrids produced in 16 factorial mating designs
between 142 Dent and 103 Flint parent lines~\cite{Westhues2017}. 
Best linear unbiased estimates (BLUEs) were computed across all factorials and 
three or more agro-ecologically diverse locations across Germany for six 
agronomically important traits by accounting for year, location, field 
replication, block and genotype effects as well as their interactions
(Table~\ref{table:TraitKey}).
All 245 parent lines have pedigree records back to the generation of their 
grandparents or deeper~\cite{Westhues2017} and were genotyped using the Illumina 
BeadChip MaizeSNP50~\cite{Ganal2011}.
After filtering for minor allele frequency ($\geq 5$\%), heterozygosity rate
($\leq 5$\%) and call frequency ($\geq 95$\%), missing genotypes were imputed 
using the Beagle software~\cite{Browning2009}.
The final number of polymorphic marker loci was 7,013 for the Dent and 6,212 for 
the Flint lines, respectively.
Gene expression data for seedlings seven days after sowing were obtained for a 
subset of 60 Dent and 43 Flint lines, that are parents of 685 hybrid progeny.
These data were obtained through two-color hybridizations using a custom 2K 
microarray (GPL22267)~\cite{Westhues2017}.
After applying established normalization procedures~\cite{Smyth2003,Ritchie2007},
BLUEs for gene expression were computed for 1,323 transcripts while adjusting
for putative effects arising from the presence of multiple factorials
(\textit{cf.} \citeNP{Westhues2017}) using the R-package
\textit{limma}~\cite{Ritchie2015a}.

@@

\Genetics2level{Population structure}
Population structure in \textit{Exp1} was examined by running a STRUCTURE
\cite{Pritchard2000} analysis with the number of putative ancestral
populations $K$ ranging from 2 to 9.
Based on the estimated probability of the data for each $K$, we chose $K=4$
and ran a principal component analysis (PCA) on the scaled and centered
genomic feature matrix using the \emph{pca} function from the \emph{R}-package
\emph{LEA} \cite{Frichot2015}.
Least-squares estimates of ancestry proportions were estimated based on sparse 
nonnegative matrix factorization algorithms~\cite{Frichot2014} using the 
\emph{snmf} function in \emph{LEA}.
Genotypes in the PCA plot (Fig.~\ref{fig:inbred-pca}) were colorized based on their
primary assignment to any of the four putative ancestral populations.
To avoid inflated predictive abilities, which can arise from large differences
in the average performance of the different subpopulations
\cite{Windhausen2012}, we reduced the panel of inbred lines to those primarily
belonging to either cluster 1 or cluster 4 in Fig.~\ref{fig:inbred-pca}.
This reduction yielded 164 genotypes, 110 of which were covered by both, genomic
and transcriptomic data.

@@

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the $l$-th group of inbred lines 
($\mathbf{W}_{l}$) has dimensions $n_{l} \times p$ where $n_{l}$ pertains to 
the number of genotypes in the $l$-th group and $p$ pertains to the number of 
features.
Note that $l$ is only defined for \textit{Exp2} where a distinction between 
the two heterotic groups is necessary whereas $l$ can be ignored for all 
models pertaining to the inbred lines of the maize diversity panel in 
\textit{Exp1}.
For the diversity panel maize lines, $n_{l}$ corresponds to the number of 
genotypes whereas, in the case of the hybrid data, $n_{l}$ corresponds to the 
number of parent lines in the $l$-th heterotic group.
In generating kernels for each predictor, all features in $\mathbf{W}_{l}$ were
centered and standardized to unit variance.
Then, each kernel can be defined as

@@

Following the notation in \citeNP{Fernando2014} the phenotypes can be untangled 
as:

@@

\subsection{Predictive ability and model validation}
For the validation of our predictions, we employed leave-one-out
cross-validation (LOOCV) routines.
In the case of the diversity panel maize inbred lines, LOOCV was performed by 
using a single genotype as a hold-out sample, which will be predicted by using
all other inbred lines for model training.
This process is repeated until all inbred lines have been used once for testing
and $n - 1$ times for model training.

@@

\begin{figure}[H]
  \centering
  \includegraphics{./tables_figures/pred_ability_inbred.pdf}
  \caption{
  Predictive abilities and SE for tropical/subtropical inbred lines from the 
  maize diversity panel and six agronomic traits.
  As predictors, genomic (G) data, transcriptomic (T) data and their 
  combination (GT) were used.
  The `Reduced' panel includes a subset of 110 lines covered by genomic and
  transcriptomic information whereas the `Full' panel comprises all 164
  genotypes.
  In the `Full' set of genotypes, transcriptomic records were imputed for 54
  lines.
  }
\label{fig:InbredResults}
\end{figure}

\Genetics2level{Exp2}
The maize hybrid data set contains information on seven agronomic traits and
three predictors, namely pedigree data (P), genomic information (G), 
transcriptomic information (T), and combinations thereof.
Pedigree records and genomic information were available for the 245 parent 
lines of all 1,521 hybrids and we denote these data as the `Full' set whereas
transcriptomic data were available for 103 parent lines of 685 hybrids, which
we denote as `Reduced'.
In the `core' set, transcriptomic data were the best single predictor for ADL,
PRO and particulaly for DMY where the predictive ability obtained with T
was 0.08 points higher than for G as the second best predictor for this trait
(Fig.~\ref{fig:HybridResults}).
In the core set, P had the highest predictive ability for STA and DMC\@.
For the FAT and SUG, G was the superior predictor inside the core set.
In the full set of 1,521 hybrids, predictive abilities obtained with G rose 
markedly for all traits with the exception of FAT\@.
The combination of genomic and transriptomic information (GT) in a single-step
model yielded a small improvement over G alone for the traits DMY and PRO\@.
Except for PRO, compared to using only G, trait prediction did not benefit from 
the single step approach when T was not superior to G in the core set.
Predictive abilities resulting from combinations of pedigree with genomic (PG)
information were better than P alone for all traits and PT improved upon P in
the `full' set for four traits.

\begin{figure}[H]
\centering
  \includegraphics{./tables_figures/pred_ability_hybrid.pdf}
  \caption{
    Predictive abilities and average, bootstrapped standard errors for the set
    of maize hybrids and six agronomic traits.
    As predictors, pedigree (P), genomic (G), transcriptomic (T) data and 
    combinations thereof we used.
    The `Full' set of genotypes comprises 1,521 hybrid progeny based on 245
    parental inbred lines, whereas the `Reduced' set comprises 685 hybrid
    progeny based on 103 parental inbred lines.
  }
\label{fig:HybridResults}
\end{figure}
